---
title: "Creating Choropleth Maps in R for Uganda"
author: "Daniel Emaasit"
date: "February 11, 2016"
output: 
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

# Choropleth Maps in R for Uganda

## What is choroplethr?
choroplethr simplifies the creation of choropleth maps in R. Choropleths are thematic maps where geographic regions, such as states, are colored according to some metric, such as the number of people who live in that state.  choroplethr simplifies this process by
    
1. Providing ready-made functions for creating choropleths using 220 different maps.
2. Providing API connections to interesting data sources for making choropleths.
3. Providing a framework for creating choropleths from arbitrary shapefiles.



## Installation

To install the latest stable release type the following from an R console:

```{r}
library(choroplethr)
library(choroplethrMaps)
data(df_pop_state)
state_choropleth(df_pop_state, title="2012 US State Population Estimates", legend="Population")
library(choroplethrMaps)

?state.regions
data(state.regions)
head(state.regions)

state_choropleth(df_pop_state,
                 title      = "2012 Population Estimates",
                 legend     = "Population",
                 num_colors = 1,
                 zoom       = c("california", "washington", "oregon"))
```

## References

Please see the following pages for more details.

1. [Introduction](http://cran.r-project.org/web/packages/choroplethr/vignettes/a-introduction.html)
1. [US State Choropleths](http://cran.r-project.org/web/packages/choroplethr/vignettes/b-state-choropleth.html)
1. [US County Choropleths](http://cran.r-project.org/web/packages/choroplethr/vignettes/c-county-choropleth.html)
1. [Country Choropleths](http://cran.r-project.org/web/packages/choroplethr/vignettes/d-country-choropleth.html)
1. [Mapping US Census Data](http://cran.r-project.org/web/packages/choroplethr/vignettes/e-mapping-us-census-data.html)
1. [Mapping World Bank WDI Data](http://cran.r-project.org/web/packages/choroplethr/vignettes/f-world-bank-data.html)
1. [Animated Choropleths](http://cran.r-project.org/web/packages/choroplethr/vignettes/g-animated-choropleths.html)
1. [Creating Your Own Maps](http://cran.r-project.org/web/packages/choroplethr/vignettes/h-creating-your-own-maps.html)
2. [Creating Administrative Level 1 Maps](http://cran.r-project.org/web/packages/choroplethr/vignettes/i-creating-admin1-maps.html)
